---
title: "Homework: Reading in Data"
author: "Firstname Lastname"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```


```{r load-sas-libraries, echo = F, message = F, warning = F}

library(SASmarkdown)
sas_enginesetup(sashtml=sashtml)

sasexe <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 75"

# Linux options (for grading, please leave this in!)
sasexe <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en"
sasopts <- "-ls 75"
```


## Instructions
Big picture: Read in the Medicare and Medicaid General Payments Data and conduct an exploratory data analysis. You should use both SAS and R to explore the data, but you might choose to use R for certain tasks and SAS for other tasks. 

When you are finished with the assignment: 
  
1. Save the file as 05_Lastname_Firstname.Rmd and compile it
2. Commit the Rmd file and corresponding html file to your homework git repo
3. Push the commit to github
4. Locate the Rmd file on github and click on the SHA link for the file
5. Paste the url of the resulting page into Canvas to submit your assignment.

Your assignment must compile on a different computer as long as the saspath is set correctly for that machine. This means you will need to use a local file path when you read the data in via R (and SAS). So when you specify your file path, it should look something like "General_Payment_Data_Sample.csv" with no "/home/xxx/Documents" or "C:\\" in front of it.  

### About the Data

To get this data, I started with the full 2018 dataset, and limited it to cash payments made to medical doctors reported in January. The full dataset is *way* too big - several GB when stored on disk. 

A sample of the full dataset is available in the template repository (and also  [here](https://github.com/srvanderplas/unl-stat850/raw/master/data/General_Payment_Data_Sample.csv)). It's 8MB after extreme trimming of the less useful columns and sampling 25% of the rows. You can read about the data [here](https://openpaymentsdata.cms.gov/about) and [here](https://www.cms.gov/openpayments/). 

### Questions to Address

- Describe the dataset and any interesting parts of the data you find. 
- What tasks are easier in R? SAS?
- What do you find that might need to be cleaned up or corrected before analysis?

You may want to include graphics (using the sample code in the book, or the [R Graph Gallery](https://www.r-graph-gallery.com/)) to show any interesting things you discover. When you include a graph, be sure to provide some contextual description of the information you want someone to take away from the graph.

## Data Exploration

### R
(You don't need to keep these headings, but I want you to have a skeleton of what the code chunks should look like)

```{r}
# Read in the CSV here

```

### SAS

Fill in your code in the chunk below, or use the generalpayment.sas file if you prefer. Write your observations in this section, referencing lines of code and/or pieces of the SAS output as necessary (do this whether or not you are using SASMarkdown). 

```{r, engine="sashtml", engine.path=sasexe, engine.opts=sasopts, collectcode = T, error = T}
/* Read in the CSV here */

```

You can delete the information below if you don't need it, but these are handy tricks for using sasmarkdown. 

```{r tricks, engine="sashtml", engine.path=sasexe, engine.opts=sasopts, error = T}
/* The collectcode=T option lets you reference data in other chunks, just like in R */
  
```
